---
title: "Causal Inference"
subtitle: "Cohorts and Risk"
author: "Ashley I Naimi"
date: "`r paste0('Spring ', format(Sys.Date(), '%Y'))`"
urlcolor: blue
output: 
    bookdown::pdf_book:
      base_format: tint::tintPdf
      toc: true
      number_sections: true
      includes:
        in_header: "../misc/preamble.tex"
      latex_engine: xelatex
    html_document:
      theme: readable
      toc: true
      toc_float: true
      number_sections: true
      css: "../misc/style.css"
bibliography: "../misc/ref_main_v4.bib"
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)

packages <- c( "data.table","tidyverse","ggplot2","ggExtra","formatR",
               "gridExtra","skimr","here","Hmisc","RColorBrewer","gmm")

for (package in packages) {
  if (!require(package, character.only=T, quietly=T)) {
    install.packages(package, repos='http://lib.stat.cmu.edu/R/CRAN')
  }
}

for (package in packages) {
  library(package, character.only=T)
}

remotes::install_github("rstudio/fontawesome")

library(fontawesome)

thm <- theme_classic() +
  theme(
    legend.position = "top",
    legend.background = element_rect(fill = "transparent", colour = NA),
    legend.key = element_rect(fill = "transparent", colour = NA)
  )
theme_set(thm)
```

\newpage
\onehalfspacing

# Complex Longitudinal Data

In this course, we will focus primarily on data where the exposure, outcome, and confounders are measured measured once. However, I will also introduce you to complex longitudinal data, and the complications that may arise when dealing with such data. For clarity, let's define complex longitudinal data. We will be dealing with data from a cohort study, individuals sampled from a well-defined target population, and clear study start and stop times (i.e., closed cohort). Data from such a cohort are **longitudinal** when they are measured repeatedly over time.^[Another such form is when data are measured repeatedly across space. We will not be dealing with these data here.]

Different scenarios can lead to longitudinal data:
\begin{itemize}
\item[1.] exposure and covariates do not vary over time, but the study outcome can occur more than once
\item[2.] exposure and covariates vary over time, but the study outcome can only occur once
\item[3.] exposure and covariates vary over time, and the study outcome can occur more than once
\end{itemize}
We will deal with data that from scenario 2 (however, it is not difficult to generalize the logic to scenario 3). 
Repeated exposure, covariate, and/or outcome measurement is what leads to "longitudinal" data. But why complex? 

Repeated measurement over time opens up the possibility of complex causal relations between past and future covariates. Suppose we measure an expsoure twice over follow-up, a covariate once, and the outcome at the end of follow-up (Figure 1). If we can assume that past exposure/covariate values do not affect future exposure/covariate values (usually a very risky assumption), we might not consider these data "complex," because we can use many standard methods we already know to analyze these data.
```{r,  out.width="10cm", fig.align='center', fig.cap="Longitudinal data that might not be considered `complex' because there is no feedback between exposure and covariates.",echo=F}
knitr::include_graphics("../figures/F1.pdf")
```
On the other hand, if past exposure/covariates affect future exposure/covariates in such a way that prior exposures or covariates confound future exposures (Figure 2), more advanced analytic techniques are needed. 
```{r,  out.width="10cm", fig.align='center', fig.cap="The simplest kind of complex longitudinal data. Note that the exposure at time zero affects the covariate at time 1 which affects the exposure at time 1. This feedback leads to confounding of the time 1 expsoure by a covariate that is affected by the prior exposure. Analysis of these data require more general methods to  account for this complex form of confounding.",echo=F}
knitr::include_graphics("../figures/F2.pdf")
```
Later, we will learn how to use methods to account for this type of complex time-varying confounding.